---
title: "Exploratory Analysis"
author: Hailey Quintavalle
format: html
---

data source: <https://www.kaggle.com/datasets/catherinerasgaitis/mxmh-survey-results?resource=download>


About the data: 

- Timestamp: date and time when form was submitted

- Age

- Primary streaming service

- Hours per day: number of hours listened to music per day

- While working: does the respondent listen to music while studying/working? 

- Instrumentalist: Does the respondent play an instrument regularly? 

- Composer: Does the respondent compose music? 

- Fav genre: favorite or top genre

- Exploratory: Does the respondent actively explore new artists/genres?

- Foreign languages: Does the respondent listen to music with lyrics in a language they are not fluent in?

- BPM: beats per minute of favorite genre 

- Frequency [Classical]: How frequently the respondent listens to classical music

...

- Anxiety: Self-reported anxiety on a scale of 0-10

- Depression: Self-reported depression, on a scale of 0-10

- Insomnia: self-reported insomnia, scale of 0-10

- OCD: self-reported obsessive compulsive disorder, 0-10

- Music effects: does music improve/worsen mental health conditions? 


```{r message = F}
library(tidyverse)
initial_data <- read_csv("data/mxmh_survey_results.csv")
```

### Cleaning up some variable names

```{r}
data <- initial_data |> rename(primary_streaming_service = `Primary streaming service`,
               hours_per_day = `Hours per day`,
               while_working = `While working`,
               fav_genre = `Fav genre`,
               foreign_languages = `Foreign languages`,
               music_effects = `Music effects`,
               Classical = `Frequency [Classical]`,
               Country = `Frequency [Country]`,
               EDM = `Frequency [EDM]`,
               Folk = `Frequency [Folk]`,
               Gospel = `Frequency [Gospel]`,
               Hip_hop = `Frequency [Hip hop]`,
               Jazz = `Frequency [Jazz]`,
               K_pop = `Frequency [K pop]`,
               Latin = `Frequency [Latin]`,
               Lofi = `Frequency [Lofi]`,
               Metal = `Frequency [Metal]`,
               Pop = `Frequency [Pop]`,
               R_B = `Frequency [R&B]`,
               Rap = `Frequency [Rap]`,
               Rock = `Frequency [Rock]`,
               Video_game_music = `Frequency [Video game music]`) |>
  select(-Permissions)
```

### Exploring Mental Health variables

```{r}
ggplot(data = data, aes(x = Anxiety)) +
  geom_histogram(fill = "orange")
```

```{r}
ggplot(data = data, aes(x = Insomnia)) +
  geom_histogram(fill = "purple4")
```


```{r}
ggplot(data = data, aes(x = Depression)) +
  geom_histogram(fill = "green4")
```

```{r}
ggplot(data = data, aes(x = OCD)) +
  geom_histogram(fill = "pink")
```



### Look into demographic around music

```{r}
data |> group_by(primary_streaming_service) |> summarise(round(mean(Age)))
```

```{r}
ggplot(data = data, aes(x = primary_streaming_service, y = Age, color = primary_streaming_service)) +
  geom_boxplot() +
  coord_flip() +
  theme(legend.position = "none")
```

```{r}
ggplot(data = data, aes(x = fav_genre, y = Age)) +
  geom_boxplot() +
  coord_flip()
```

```{r}
ggplot(data = data, aes(x = hours_per_day, y = Age)) +
  geom_point()
```

```{r}
ggplot(data = data, aes(x = while_working, y = Age)) +
  geom_boxplot() +
  coord_flip()
```

## Recode music effects variable into numeric values

**should recode all music genre variables into numeric variables 

```{r}
data |> mutate(music_effects1 = if_else(music_effects == "Improve", 1, 0))
```


## Brainstorm ideas

Possible questions:

- **Primary Question:** which musical genres seem to `help` the most for different health disorders?

- 

## Hours listening per day and mental health

```{r}
mod_1 <- lm(hours_per_day ~ Anxiety + Depression + OCD + Insomnia, data = data)
summary(mod_1)
```

## Modeling genre with Anxiety

```{r}
mod_a <- glm(Anxiety ~ Classical, data = data)
summary(mod_a)
```





















