---
title: "DATA 334 Final Project"
format: gfm
---



# **Music and Mental Health!**

This project aims to explore the frequency that individuals struggling with anxiety, depression, OCD, or insomnia listen to specific music genres, and to determine if those who listen believe music had effected their overall mental health. 

## Data

The data used in this project comes from an undergraduate student from the University of Washington, uploaded to [Kaggle](https://www.kaggle.com/datasets/catherinerasgaitis/mxmh-survey-results?resource=download). 

The online survey obtained responses from 736 people, gathering information about whether music has improved or worsened the state of individuals’ mental health. Within the survey, people were asked to rate how much they listen to 16 different music genres and were also presented with symptoms of Anxiety, Depression, OCD, and Insomnia and asked to rank each disorder on a scale of 0 to 10; 0 representing not having experienced this disorder at all and 10 representing having experienced this disorder regularly or to an extreme.

## Results

A Shiny App was created to aid in the exploration of responses, allowing users to choose genre, disorder, minimum number of symptoms reported, and the overall effect music has had on metal health. 

The app produces a number of plots similar to the one below. This plot contains the reported frequency of listening to **rock** music of those who have experienced **anxiety** symptoms and say music has **improved** their mental health.

```{r echo = F, message = F, warning = F}
library(tidyverse)
library(pander)
library(plotly)

initial_data <- read_csv("data/mxmh_survey_results.csv")
wrangled_data <- initial_data |> rename(primary_streaming_service = `Primary streaming service`,
               hours_per_day = `Hours per day`,
               while_working = `While working`,
               fav_genre = `Fav genre`,
               foreign_languages = `Foreign languages`,
               music_effects = `Music effects`,
               Classical = `Frequency [Classical]`,
               Country = `Frequency [Country]`,
               EDM = `Frequency [EDM]`,
               Folk = `Frequency [Folk]`,
               Gospel = `Frequency [Gospel]`,
               `Hip Hop` = `Frequency [Hip hop]`,
               Jazz = `Frequency [Jazz]`,
               `K-Pop` = `Frequency [K pop]`,
               Latin = `Frequency [Latin]`,
               Lofi = `Frequency [Lofi]`,
               Metal = `Frequency [Metal]`,
               Pop = `Frequency [Pop]`,
               `R&B` = `Frequency [R&B]`,
               Rap = `Frequency [Rap]`,
               Rock = `Frequency [Rock]`,
               `Video Game` = `Frequency [Video game music]`) |>
  select(-Permissions) |> 
  mutate(music_improve = recode(music_effects, 
                                'Improve'= '1', 
                                'No effect'='0', 
                                'Worsen' = '-1')) |>
  mutate(music_improve = as.numeric(music_improve))
```


```{r echo = F}
wrangled_data |> pivot_longer(c(Classical, Country, EDM, Folk, Gospel, `Hip Hop`, Jazz, 
                                'K-Pop', Latin, Lofi, Metal, Pop, 'R&B', Rap, Rock, 'Video Game'),
                              names_to = "genres", 
                              values_to = "genre_listen") |>
  pivot_longer(c(Anxiety, Depression, Insomnia, OCD), 
               names_to = "disorder", 
               values_to = "n_symptoms") |>
  mutate(n_symptoms = as.numeric(n_symptoms)) |>
  filter(music_effects == "Improve",
         disorder == "Anxiety",
         genres == "Rock",
         n_symptoms >= 1) |>
  group_by(genre_listen) |>
  summarise(n = n()) |>
  ggplot(aes(x = genre_listen, y = n)) +
  geom_col(aes(fill = genre_listen)) +
  scale_fill_viridis_d() +
  theme_bw(base_size = 16) +
  theme(legend.position = "none",
        axis.title.x = element_text(vjust = -1)) +
  labs(x = "How often do you listen to rock music?",
       y = "Number of Responses",
       title = "Reported Music IMPROVED Mental Health")
```

Findings suggest that people frequently listening to rock music report music improving their mental health, regardless of reported symptoms. Further exploration of the Shiny App also reveals more individuals who listen to video game music frequently *and* have experienced depression report music worsening their mental health.

It is important to note that these conclusions are drawn from self-reported data, which is prone to be effected by personal biases and uncontrollable outside variables.


